|

### 2025-11-27 manuscript/pre_print.qmd
Abstract and discussion edits (for some reason didn't previously save)

---
title: |
  Resting energy expenditure of women with and without polycystic ovary syndrome: a systematic review and meta-analysis
date: last-modified
date-format: long
format:
  pdf:
    # documentclass: article
    pdf-engine: xelatex
    mainfont: "Latin Modern Roman"
    include-in-header:
      text: |
        \usepackage{lineno}\linenumbers
        \usepackage{setspace}\doublespacing
        \usepackage[font=scriptsize]{caption}
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
geometry:
  - left=1in
  - right=1in
  - top=1in
  - bottom=1in
author:
  - name: Richard Kirwan
    affiliations:
      - ref: 1
    orcid: 0000-0003-4645-0077
    corresponding: true
    email: R.P.Kirwan@ljmu.ac.uk
  - name: Leigh Peele
    affiliations:
      - ref: 2
    email: leigh@macrofactorapp.com
  - name: Greg Nuckols
    affiliations:
      - ref: 2
    orcid: 0009-0004-2679-7967
    email: greg@macrofactorapp.com
  - name: Georgia Kohlhoff
    affiliations:
      - ref: 3
    # orcid: 0000-0002-8003-0757
    email: g_kohlhoff@hotmail.co.uk
  - name: Hannah Cabré
    affiliations:
      - ref: 4
    orcid: 0000-0002-3235-6867
    email: Hannah.Cabre@pbrc.edu
  - name: Alyssa Olenick
    affiliations:
      - ref: 5
    orcid: 0000-0002-2217-8067
    corresponding: true
    email: alyssaolenickphd@gmail.com
  - name: James Steele
    affiliations:
      - ref: 6
      - ref: 2
      - ref: 7
    orcid: 0000-0002-8003-0757
    corresponding: true
    email: james@steele-research.com
affiliations:
  - id: 1
    name: Research Institute for Sports and Exercise Sciences, Liverpool John Moores Universtiy, Liverpool, UK
  - id: 2
    name: MacroFactor, Stronger by Science LLC, Raleigh, North Carolina, USA
  - id: 3
    name: Flourishing Health, Liverpool, UK
  - id: 4
    name: Reproductive Endocrinology and Women’s Health Research Program, Pennington Biomedical Research Center, Baton Rouge, Louisiana, USA
  - id: 5
    name: Dr Alyssa Olenick LLC
  - id: 6
    name: Steele Research Limited, Eastleigh, Hampshire, UK
  - id: 7
    name: School of Health and Biomedical Sciences, Royal Melbourne Institute of Technology, Melbourne, VIC, Australia
abstract: |
  Background: Polycystic ovary syndrome (PCOS) is the most common endocrine condition affecting women of reproductive age. Women who are living with PCOS are also more likely to fall into overweight or obese body mass index (BMI) categories and this is assumed to be a result of a lower resting energy expenditure (REE). The assumption that REE is lower in women living with PCOS may also influence lifestyle-based interventions in this group. Evidence supporting this reduced REE compared to women without PCOS, however, is not consistent. Objectives: To systematically search and meta-analyse the existing literature to estimate and describe the difference in REE between women with and without PCOS. Design: A systematic search was conducted using PubMed, Medline and Web of Science databases of published research from January 1990 to January 2025. Studies that measured REE in women living with PCOS, both with and without control arms of women without PCOS, were included. Results: Seventeen independent studies were identified. Study populations ranged from X to X participants, with mean ages ranging from X to X years. Conclusions: TO ADD
keywords: Polycystic ovary syndrome, PCOS, resting energy expenditure, REE, basal metabolic rate, BMR
bibliography: 
  mylibrary.bib
csl: apa.csl
execute: 
  echo: false
  message: false
  warning: false
---
```{r}
library(tidyverse)
library(flextable)
library(tidybayes)

@@

# Methods
This systematic review and meta-analysis was pre-registered on PROSPERO ([CRD42024601434](https://www.crd.york.ac.uk/PROSPERO/view/CRD42024601434)) initially on the $3^{rd}$ of December 2024 and performed in accordance with the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) statement guidelines [@pagePRISMA2020Statement2021a]. The primary aim of this review was to examine the descriptive question "Does resting energy expenditure (REE) differ between women with and without polycystic ovary syndrome (PCOS)?". We summarise and describe the studies in addition to quantitatively synthesising their results via meta-analysis.

@@

## Data extraction (selection and coding)
Bibliometric data including authors, journal, and article titles were extracted. Descriptive statistics for age, body mass, fat mass, fat free mass, height, BMI, race, physical activity levels, country of investigation, information regarding glucose/insulin regulation and insulin resistance status (where available), diagnostic criteria for PCOS, and measurement method and device were extracted for each arm within each study in addition to sample size. Descriptive characteristics were then tabulated across studies for reporting.

@@

This study sought to estimate and describe the magnitude of difference in REE between women with and without PCOS. Most studies identified in the systematic review, and included in the meta-analysis, used indirect calorimetry as the primary measure of REE and assessed women with PCOS who were insulin resistant and categorised as being in overweight or obese BMI categories compared to healthy controls. Our results indicate there is only a small magnitude of difference in REE (`r round(main_arm_mean_effects_contrast_condition$draw)` kcal/day [95% quantile interval: `r round(main_arm_mean_effects_contrast_condition$.lower)` kcal/day to `r round(main_arm_mean_effects_contrast_condition$.upper)`kcal/day]) between women with PCOS and those without and also little difference in between person variation between the groups based on the ratio of standard deviations (`r round(main_arm_variance_effects_contrast_condition$draw,2)` [95% quantile interval: `r round(main_arm_variance_effects_contrast_condition$.lower,2)` to `r round(main_arm_variance_effects_contrast_condition$.upper,2)`]). These findings challenge the widely held belief that PCOS is inherently associated with a slower metabolism (PMID: 32244780), predisposing women with PCOS to weight gain. This belief may mistakenly lead to recommendations centred on slightly more severe calorie restriction to achieve weight-loss goals, compared with recommendations for the general population, as primary management strategies for women with PCOS [@ozgensaydamWeightManagementStrategies2021]. Recognising that there may be minimal differences in REE between women with PCOS and those without can inform both clinical and public practices, potentially leading to a shift in focus away from a requirement for more severe caloric restriction as a primary method of treatment towards more comprehensive, individualised, and psychologically safe approaches to care [@teedeRecommendations2023International2023].

The current study suggests that REE may not be the primary barrier to weight regulation in PCOS given small group-level differences between women with PCOS and those without (1442 kcal/day vs. 1472 kcal/ day, respectively). Similar to our findings, a case-control study comparing women with PCOS to healthy controls demonstrated REE was similar between groups (PMID: 38867472); however, REE adjusted for fat-free mass was lower in women with PCOS suggesting the importance of fat-free tissue in energy regulation. Indeed, a recent systematic review of mechanisms for metabolic dysfunction found excess androgen drives metabolic issues within adipose tissue and muscle tissue contributing to complications like obesity and insulin resistance (PMID: 32244780). Taken together, these findings highlight that factors beyond REE, such as hormonal and tissue-specific metabolic effects, likely play a more significant role in weight regulation challenges in women with PCOS.

Women with PCOS are routinely encouraged to participate in caloric restriction for weight management, advice that may inadvertently foster disordered eating. The prevalence of any eating disorder in women with PCOS has been reported at up to 62%, with affected women three to six times more likely than controls to experience an eating disorder (PMID: 38408541). Specific conditions such as binge-eating disorder (about 17%), bulimia nervosa (about 6 per cent) and night-eating syndrome (about 13%) are all markedly more common than in the general female population (PMID: 28104244, PMID: 27244869). Approximately 60% of women with PCOS report binge or comfort eating, with 39% displaying clinically significant behaviours (PMID: 27825940). 
The pathways linking PCOS and disordered eating are multifactorial. Biological mechanisms such as hyperandrogenism, hyperinsulinaemia, and altered ghrelin and leptin signalling can heighten hunger, carbohydrate cravings, and appetite variability (PMID: 21421016). Frequent hypoglycaemia and associated mood changes have also been observed, which can trigger compensatory eating or binge episodes (PMID: 21421016). These physiological processes interact with psychological and social stressors, including infertility concerns, conflicting nutrition advice, chronic dieting, and exposure to idealised body images on social media, which together compound vulnerability to disordered eating (PMID 33153014). Moreover, eating disorders themselves can disrupt endocrine function, potentially worsening PCOS symptoms and creating a self-reinforcing cycle (PMID 39115340, PMID 27244869). Understanding these intertwined biological, psychological, and social influences underscores the importance of avoiding overly restrictive dietary advice that can exacerbate feelings of failure, hunger dysregulation, and shame (PMID31615157). When metabolic rate is not impaired, recommending large caloric deficits is not only unnecessary but may be actively harmful.
A further barrier to effective management is weight stigma, which remains pervasive in both healthcare and everyday life (PMID: 39462252). Women with PCOS often report being labelled as “non-compliant” or “lacking willpower” when they do not lose weight rapidly, despite evidence of complex metabolic and psychosocial contributors (PMID: 37580314). Such attitudes are not benign. Weight stigma has been linked to poorer self-esteem, depression, anxiety, avoidance of medical care, and increased disordered eating. Physiologically, stigma-related stress elevates chronic cortisol, blood pressure, and insulin resistance, thereby worsening cardiometabolic outcomes (PMID: 37580314). Repeated cycles of weight loss and regain, driven by shame and unrealistic expectations, are associated with higher long-term weight, adverse lipid profiles, and elevated cardiovascular risk (PMID 37436934, PMID 23046078, PMID 10792339). Encouragingly, the 2023 international PCOS guidelines now recognise weight stigma as a determinant of health and call for its reduction across clinical and public health settings. Addressing stigma is therefore both an ethical imperative and a practical strategy to improve health outcomes.
Given these insights, the finding of a similar REE  in women with PCOS compared to those without PCOS invites a paradigm shift towards weight-neutral, individualised, and empowering care. Without evidence of metabolic disadvantage, interventions need not centre on extreme energy deficits. Instead, practitioners can focus on improving insulin sensitivity, inflammatory markers, and hormonal regulation through balanced nutrition, physical activity for enjoyment, adequate sleep, and stress management. Screening for disordered eating and providing evidence-based psychological support should be standard components of PCOS care. Approaches grounded in intuitive or mindful eating encourage self-trust and attunement to hunger and fullness cues, while dismantling moral judgements about food. This can stabilise glycaemic control, reduce binge episodes, and promote sustainable behaviour change (PMID: 32006391). Health coaching and dietetic practice should also include ongoing reflection on personal biases, the use of non-stigmatising language, and the creation of inclusive environments for people in all body sizes.
Importantly, this reframing is empowering for women with PCOS. Understanding that they are not “metabolically disadvantaged” enables them to move forward in their health journey without the burden of restrictive calorie targets or the belief that they are inherently at fault. By shifting attention towards metabolic-health parameters such as insulin resistance, stress management, and healthy lifestyle, women can pursue improvements in energy, mood, fertility, and quality of life on their own terms. When the focus becomes nourishment rather than restriction, self-care rather than self-control, and progress rather than perfection, sustainable wellbeing becomes achievable.

Strengths and Limitations
The current study has multiple strengths stemming from its preregistered, comprehensive methodology and Bayesian approach. This statistical framework allowed us to incorporate studies with and without control groups to better estimate mean REE in women with and without PCOS and to perform multiple sensitivity analyses that confirmed the stability of our findings. However, a limitation here is that variability in methods across studies, such as differences in PCOS diagnostic criteria and REE testing protocols, may have influenced results. Furthermore, some studies controlled for body weight or body composition when reporting REE values, while others did not. We accounted for this by estimating or converting reported data to obtain unadjusted REE values across all groups, thereby reducing this variability. Another limitation is that, due to fewer total control groups than PCOS groups, informed priors were required in several statistical models. Additionally, some studies reported data inconsistencies that could not be clarified (e.g., Georgopoulos et al.) and were accounted for in our analysis. Finally, most included studies were cross-sectional or baseline assessments within intervention trials, which limits causal inference.

Conclusion
In conclusion, the findings from this meta-analysis indicate that resting energy expenditure does not meaningfully differ between women with and without PCOS. Group-level differences in resting energy expenditure were small, insignificant, or not physiologically relevant. Additionally, variability in resting energy expenditure was also similar, suggesting that a lower baseline metabolic rate is not a causal factor for the weight-related challenges often associated with PCOS. These findings challenge the popular narrative that women with PCOS have a lower resting metabolism and may help better inform dietary interventions and nutritional support for these individuals. Future research should include more standardized REE measurement and reporting protocols, greater data transparency, consistent control and reporting of body weight or body composition, the presentation of both absolute and relative REE, and more precise characterization of PCOS phenotypes. Overall, these findings support the conclusion that PCOS does not negatively influence resting energy expenditure and may help practitioners and researchers focus on targeted lifestyle interventions rather than negatively framed metabolic assumptions.

>>>>

---
title: |
  Resting energy expenditure of women with and without polycystic ovary syndrome: a systematic review and meta-analysis
date: last-modified
date-format: long
format:
  pdf:
    # documentclass: article
    pdf-engine: xelatex
    mainfont: "Latin Modern Roman"
    include-in-header:
      text: |
        \usepackage{lineno}\linenumbers
        \usepackage{setspace}\doublespacing
        \usepackage[font=scriptsize]{caption}
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
geometry:
  - left=1in
  - right=1in
  - top=1in
  - bottom=1in
author:
  - name: Richard Kirwan
    affiliations:
      - ref: 1
    orcid: 0000-0003-4645-0077
    corresponding: true
    email: R.P.Kirwan@ljmu.ac.uk
  - name: Leigh Peele
    affiliations:
      - ref: 2
    email: leigh@macrofactorapp.com
  - name: Greg Nuckols
    affiliations:
      - ref: 2
    orcid: 0009-0004-2679-7967
    email: greg@macrofactorapp.com
  - name: Georgia Kohlhoff
    affiliations:
      - ref: 3
    # orcid: 0000-0002-8003-0757
    email: g_kohlhoff@hotmail.co.uk
  - name: Hannah Cabré
    affiliations:
      - ref: 4
    orcid: 0000-0002-3235-6867
    email: Hannah.Cabre@pbrc.edu
  - name: Alyssa Olenick
    affiliations:
      - ref: 5
    orcid: 0000-0002-2217-8067
    corresponding: true
    email: alyssaolenickphd@gmail.com
  - name: James Steele
    affiliations:
      - ref: 6
      - ref: 2
      - ref: 7
    orcid: 0000-0002-8003-0757
    corresponding: true
    email: james@steele-research.com
affiliations:
  - id: 1
    name: Research Institute for Sports and Exercise Sciences, Liverpool John Moores Universtiy, Liverpool, UK
  - id: 2
    name: MacroFactor, Stronger by Science LLC, Raleigh, North Carolina, USA
  - id: 3
    name: Flourishing Health, Liverpool, UK
  - id: 4
    name: Reproductive Endocrinology and Women’s Health Research Program, Pennington Biomedical Research Center, Baton Rouge, Louisiana, USA
  - id: 5
    name: Dr Alyssa Olenick LLC
  - id: 6
    name: Steele Research Limited, Eastleigh, Hampshire, UK
  - id: 7
    name: School of Health and Biomedical Sciences, Royal Melbourne Institute of Technology, Melbourne, VIC, Australia
abstract: |
  Context: Polycystic ovary syndrome (PCOS) is common in reproductive-age women, who often have higher BMI classification. This is assumed to stem from lower resting energy expenditure (REE), influencing lifestyle intervention guidelines. However, evidence for reduced REE in women with PCOS compared with those without is inconsistent. Objective: To systematically search and meta-analyse the existing literature to estimate and describe the difference in REE between women with and without PCOS. Data Sources: A systematic search was conducted using PubMed, Medline and Web of Science databases of published research from January 1990 to January 2025. Study Selection: Studies that measured REE in women living with PCOS, both with and without control arms of women without PCOS, were included. Data Extraction: Bibliometric, demographic, and REE data was extracted by one investigator and checked in triplicate. Data Synthesis: Thirteen studies were included in a Bayesian arm-based multiple condition comparison (i.e., network) type meta-analysis model with informative priors was used to compare both mean REE, and between person variation in REE, between women with and without PCOS. Mean REE differed between groups by 30 kcal/day [95% quantile interval: -47 to 113 kcal/day] and the contrast ratio for between person standard deviations was 0.98 [95% quantile interval: 0.71 to 1.33]. Conclusions: These findings indicate that REE does not meaningfully differ between women with and without PCOS. Group-level differences in resting energy expenditure are small, insignificant, or not physiologically relevant.
keywords: Polycystic ovary syndrome, PCOS, resting energy expenditure, REE, basal metabolic rate, BMR
bibliography: 
  mylibrary.bib
csl: apa.csl
execute: 
  echo: false
  message: false
  warning: false
---
library(tidyverse)
library(flextable)
library(tidybayes)

@@

# Materials and Methods
This systematic review and meta-analysis was pre-registered on PROSPERO ([CRD42024601434](https://www.crd.york.ac.uk/PROSPERO/view/CRD42024601434)) initially on the $3^{rd}$ of December 2024 and performed in accordance with the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) statement guidelines [@pagePRISMA2020Statement2021a]. The primary aim of this review was to examine the descriptive question "Does resting energy expenditure (REE) differ between women with and without polycystic ovary syndrome (PCOS)?". We summarise and describe the studies in addition to quantitatively synthesising their results via meta-analysis.

@@

## Data extraction (selection and coding)
Data was extracted by one investigator and checked in triplicate. Bibliometric data including authors, journal, and article titles were extracted. Descriptive statistics for age, body mass, fat mass, fat free mass, height, BMI, race, physical activity levels, country of investigation, information regarding glucose/insulin regulation and insulin resistance status (where available), diagnostic criteria for PCOS, and measurement method and device were extracted for each arm within each study in addition to sample size. Descriptive characteristics were then tabulated across studies for reporting.

@@

This study sought to estimate and describe the magnitude of difference in REE between women with and without PCOS. Most studies identified in the systematic review, and included in the meta-analysis, used indirect calorimetry as the primary measure of REE and assessed women with PCOS who were insulin resistant and categorised as being in overweight or obese BMI categories compared to healthy controls. Our results indicate there is only a small magnitude of difference in REE (`r round(main_arm_mean_effects_contrast_condition$draw)` kcal/day [95%QI: `r round(main_arm_mean_effects_contrast_condition$.lower)` kcal/day to `r round(main_arm_mean_effects_contrast_condition$.upper)`kcal/day]) between women with PCOS and those without. Further, there is little difference in between person variation between the groups based on the ratio of standard deviations (`r round(main_arm_variance_effects_contrast_condition$draw,2)` [95%QI: `r round(main_arm_variance_effects_contrast_condition$.lower,2)` to `r round(main_arm_variance_effects_contrast_condition$.upper,2)`]) suggesting that, despite individual differences in REE, PCOS is not systematically associated with lesser or greater individual variability. These findings challenge the widely held belief that PCOS is inherently associated with a slower metabolism (PMID: 32244780), predisposing women with PCOS to weight gain. This belief may mistakenly lead to recommendations centred on slightly more severe calorie restriction to achieve weight-loss goals, compared with recommendations for the general population, as primary management strategies for women with PCOS [@ozgensaydamWeightManagementStrategies2021]. Recognising that there may be minimal differences in REE between women with PCOS and those without can inform both clinical and public practices, potentially leading to a shift in focus away from a requirement for more severe caloric restriction as a primary method of treatment towards more comprehensive, individualised, and psychologically safe approaches to care [@teedeRecommendations2023International2023].

The current study suggests that REE may not a barrier to weight regulation in PCOS given small group-level differences between women with PCOS and those without (`r round(main_arm_variance_effects_preds_condition$draw[2])` kcal/day [95%QI:`r round(main_arm_variance_effects_preds_condition$.lower[2])` kcal/day to `r round(main_arm_variance_effects_preds_condition$.upper[2])` kcal/day] versus `r round(main_arm_variance_effects_preds_condition$draw[1])` kcal/day [95%QI:`r round(main_arm_variance_effects_preds_condition$.lower[1])` kcal/day to `r round(main_arm_variance_effects_preds_condition$.upper[1])` kcal/day], respectively). If anything, REE may be slightly higher in women with PCOS compared to healthy women without PCOS. BMI of women with PCOS ranged from `r summary_descriptives$min_bmi[2]` to `r summary_descriptives$max_bmi[2]` was typically greater than women without PCOS which ranged `r summary_descriptives$min_bmi[1]` to `r summary_descriptives$max_bmi[1]` and this may explain the slightly greater REE in the former group. However, for those studies where we could extract or estimate BMI, we fit an additional exploratory model extending the main arm based model and extracted an estimate of the difference adjusted for BMI as a fixed effect. Adjusted to the median value of average BMI of the control arms of women without PCOS (BMI = `r main_arm_mean_effects_contrast_condition_bmi$m_bmi` kg/$\textrm{m}^2$) the difference in REE between was (`r round(main_arm_mean_effects_contrast_condition_bmi$estimate)` kcal/day [95%QI: `r round(main_arm_mean_effects_contrast_condition_bmi$conf.low)` kcal/day to `r round(main_arm_mean_effects_contrast_condition_bmi$conf.high)`kcal/day]) between women with PCOS and those without. However, one of the studies included in our analysis [@tosiRestingEnergyExpenditure2024] reported that, whilst there was little difference in unadjusted REE, when REE was adjusted for fat-free mass it was lower in women with PCOS suggesting the potential importance of fat-free tissue in energy regulation. Yet, similarly to the additional exploratory model including BMI, when we adjusted for fat-free mass including those studies reporting this characteristic (fat-free mass = `r main_arm_mean_effects_contrast_condition_fat_free_mass$m_fat_free_mass` kg) the difference in REE between was (`r round(main_arm_mean_effects_contrast_condition_fat_free_mass$estimate)` kcal/day [95%QI: `r round(main_arm_mean_effects_contrast_condition_fat_free_mass$conf.low)` kcal/day to `r round(main_arm_mean_effects_contrast_condition_fat_free_mass$conf.high)`kcal/day]). As such, even adjusted for both BMI and fat-free mass, there seems to be little difference in REE between women with, and without, PCOS. Yet, a recent systematic review of mechanisms for metabolic dysfunction has reported excess androgen drives metabolic issues within adipose tissue and muscle tissue contributing to complications like obesity and insulin resistance (PMID: 32244780). Taken together, these findings highlight that factors beyond REE and typical correlates of this including BMI or fat-free mass, such as hormonal and tissue-specific metabolic effects, may play a more significant role in weight regulation challenges in women with PCOS.

As noted, women with PCOS are more likely to engage in weight-loss attempts [@pesonenPolycysticOvarySyndrome2023] and there could be concerns that this could inadvertently further foster the already well documented disordered eating in this population [@lalonde-besterPrevalenceEtiologyEating2024; @jeanesBingeEatingBehaviours2017; @leeIncreasedRiskDisordered2017; @bernadettPrevalenceEatingDisorders2016]. The pathways linking PCOS and disordered eating are multifactorial. Biological mechanisms such as hyperandrogenism, hyperinsulinaemia, and altered ghrelin and leptin signalling can heighten hunger, carbohydrate cravings, and appetite variability [@barryImpactEatingBehavior2011]. Frequent hypoglycaemia and associated mood changes have also been observed, which can trigger compensatory eating or binge episodes [@barryImpactEatingBehavior2011]. These physiological processes interact with psychological and social stressors, including infertility concerns, conflicting nutrition advice, chronic dieting, and exposure to idealised body images on social media, which together compound vulnerability to disordered eating [@steegers-theunissenPolycysticOvarySyndrome2020]. Moreover, eating disorders themselves can disrupt endocrine function, potentially worsening PCOS symptoms and creating a self-reinforcing cycle [@cooneyIncreasedPrevalenceBinge2024; @bernadettPrevalenceEatingDisorders2016].

Understanding the intertwined biological, psychological, and social influences suggests the importance of considering whether restrictive dietary advice is appropriate given it may exacerbate feelings of failure, hunger dysregulation, and shame [@pirottaDisorderedEatingBehaviours2019]. These concerns, coupled with the lack of difference in REE between women with and without PCOS might suggest that energy restriction based dietary interventions for weight-loss may be unnecessary. But, there is also evidence supporting the effect of energy restricted dietary interventions for improving PCOS symptoms [@aleneziImpactDietInducedWeight2024; @holteRestoredInsulinSensitivity1995] and they are recommended in international guidelines [@teedeRecommendations2023International2023]. Encouragingly though, these guidelines also recognise weight stigma as a determinant of health and call for its reduction across clinical and public health settings. Evidently the greater prevalence of women with PCOS falling into overweight and obese BMI categories compared to women without PCOS [@barberWhyAreWomen2022; @barberObesityPolycysticOvary2021] is unlikely to be due to differences in REE and so, despite the potential effectiveness of energy restriction dietary interventions, there is potential value in moving towards more weight-neutral, individualised, and empowering care following holistic guidelines recommending multiple approaches to management [@teedeRecommendations2023International2023; @ozgensaydamWeightManagementStrategies2021].

## Strengths and Limitations
The current study has multiple strengths stemming from its preregistered, comprehensive methodology and Bayesian statistical approach. This statistical framework allowed us to incorporate studies with and without control groups to better estimate REE in women with and without PCOS and to perform multiple sensitivity analyses that confirmed the stability of our findings. However, a limitation here is that variability in methods across studies, such as differences in PCOS diagnostic criteria and REE testing protocols, may have influenced results and the relatively small number of studies overall limits the extent to which we can explore these potential moderators. Furthermore, some studies controlled for body weight or body composition when reporting REE values, while others did not. We accounted for this by estimating or converting reported data to obtain unadjusted REE values across all groups, thereby reducing this variability and further as noted above provided estimates adjusted for BMI and fat-free mass both of which had little influence on our conclusions. Another limitation is that, due to fewer total control groups than PCOS groups, informed priors were required in several statistical models. However, in the context of Bayesian meta-analysis this can also be considered a strength. Additionally, some studies reported data inconsistencies that could not be clarified (e.g., [@kritikouA2BV3Adrenergic2006; @saltamavrosAlpha2Beta2007; @koikaAssociationPro12AlaPolymorphism2009; @georgopoulosBasalMetabolicRate2009]) and were dropped from our main analysis though our conclusions again did not qualitative change when we conducted sensitivity analyses including these studies. Finally, most included studies were cross-sectional or baseline assessments within intervention trials, which limits causal inference. Indeed, we did not pre-register any kind of causal model (e.g., a directed acyclic graph) to inform our analysis approach for causal inference and as such have been explicit about the estimates presented as being descriptive.

# Conclusion
In conclusion, the findings from this meta-analysis indicate that REE does not meaningfully differ between women with and without PCOS. Group-level differences in REE were small, insignificant, or not physiologically relevant. Additionally, variability in REE between individuals was also similar. These results suggest that a lower baseline REE is not associated with the weight-related challenges often associated with PCOS. These findings challenge the popular narrative that women with PCOS have a lower REE and may help better inform dietary interventions and nutritional support for these individuals. Future research should include more standardized REE measurement and reporting protocols, greater data transparency, consistent control and reporting of body weight or body composition, the presentation of both absolute and relative REE, and more precise characterization of PCOS phenotypes. Overall, these findings support the conclusion that PCOS is not negatively associated with REE and may help practitioners and researchers focus on individually targeted and holistic lifestyle interventions rather than negatively framed interventions based on unsupported assumptions regarding REE.

#
